Seeing Double: Transient Dips and Photometric Binarity in PTFO$\,$8-8695

### 2020-04-20 paper/ms.tex
add words and figures on astrometric binarity

\documentclass[12pt,twocolumn,tighten]{aastex62}
%\documentclass[12pt,twocolumn,tighten,trackchanges]{aastex62}
\usepackage{amsmath,amstext,amssymb}
\usepackage[T1]{fontenc}
\usepackage{apjfonts}
\usepackage[figure,figure*]{hypcap}
\usepackage{graphics,graphicx}
\usepackage{hyperref}
\usepackage{natbib}

@@

@@

\begin{abstract}
  PTFO$\,$8-8695b is a candidate hot Jupiter in the 7--10 million year
  old Orion-OB1a cluster. We inspected data from TESS and Gaia to
  clarify whether it is actually a planet.  The Gaia data show that
  PTFO$\,$8-8695 is a photometric binary with respect to members of
  its kinematic group.  The TESS lightcurve shows that the dominant
  variability in this system is a sinusoid with a ``long'' period
  $P_{\rm \ell}=11.96\,$hr, presumably caused by stellar rotation.
  Also present is a complex signal, previously identified as the
  planet candidate, that repeats with a ``short'' period $P_{\rm s}=
  10.74\,$hr.  The two signals beat every 4.48 days.  Although there
  is a dip in the short-period signal, ground-based photometry from
  the past decade shows that the orbital phase of the dip seems to
  have instantaneously jumped, at least once, and perhaps twice.
  Planets do not ``jump'' in orbital phase.  Given the evidence, we
  believe that PTFO$\,$8-8695 is a binary M dwarf in which one star
  shows the long rotation signal, and the other star is showing
  ``transient dipping'' that has been seen in a few other young
  weak-lined T Tauri stars.  The origin of these transient dips is
  still undetermined, but our preferred explanation is eclipses of
  clouds of gas or dust at the Keplerian co-rotation radius.
\end{abstract}

\object{PTFO 8-8695}
\object{PTFO 8-8695b}
\object{CVSO 30}

@@

Alternatively, between 0.1\% and 1\% of rapidly rotating low-mass
stars in $\mathcal{O}$(10)$\,$Myr old associations show narrow dips in
phase with strong stellar rotation signals \citep{rebull_usco_2018}.
The dips can persist over months, but their depths often vary, and
sometimes change immediately after stellar flares.  The explanation
proposed by {\it e.g.}, \citet{stauffer_orbiting_2017} and
\citet{david_transient_2017} is that a circumstellar cloud of dust or
gas might be orbiting near the Keplerian co-rotation radius.  For this
explanation to be viable for \ptfo, a clear determination of the
stellar rotation period is necessary.  To date, this has not been
possible
\citep{van_eyken_ptf_2012,koen_multicolour_2015,raetz_yeti_2016}.

We begin in Section~\ref{sec:observations} by describing newly
available observations from the Transiting Exoplanet Survey Satellite
(TESS; \citealt{ricker_transiting_2015}) and Gaia
\citep{gaia_collaboration_gaia_2018}.  The TESS lightcurve shows two
distinct signals, which we extract and analyze in
Section~\ref{sec:tess}.  A long-period sinusoid repeats every 12.0
hours, and is probably stellar rotation.  A short-period ``dip +
complex modulation'' repeats every 10.7 hours.  Analyzing the Gaia
data in Section~\ref{sec:gaia}, we show that relative to its kinematic
group, \ptfo\ is a photometric binary.  We collect and discuss the
puzzle pieces in Section~\ref{sec:discussion}.  The orbital phase of
the dip seems to have instantaneously changed over the past decade.
In addition, a number of other young stars show similar lightcurve
morphologies.  We therefore argue that PTFO$\,$8-8695 is a binary M
dwarf in which one star shows a rotation signal, and the other is
showing ``transient dipping'' that is likely caused by eclipses of
clouds of material at the Keplerian co-rotation radius.
Section~\ref{sec:conclusions} summarizes our main points.

@@

The 2-minute stacks for \ptfo\ were then reduced to lightcurves by the
Science Processing Operations Center (SPOC) at NASA
Ames~\citep{jenkins_tess_2016}.  Our main analysis used the resulting
Presearch Data Conditioning (PDC) lightcurve.  The PDC lightcurve
aperture used pixels chosen to maximize the SNR of the total flux of
the target \citep{smith_kepler_apertures_2017}.  Non-astrophysical
variability was removed through the methods discussed by
\citet{smith_kepler_PDC_2017}.

As an independent check on the shorter cadence SPOC light-curve, we
separately processed the 30-minute image stacks as part of the Cluster
Difference Imaging Photometric Survey (CDIPS;
\citealt{bouma_cluster_2019}).  The CDIPS lightcurve used a circular
aperture with radius 1 pixel.

To clean the data, we removed all points with non-zero quality flags
\citep[{\it e.g.},][]{tess_data_product_description_2018}.  We also
masked out the first and last 6 hours of each orbit, since there is
often systematic red noise during those times.  Both the CDIPS and PDC
lightcurves showed a clear discontinuous ``jump'' in the last few days
of orbit 20, which seemed likely to be an instrumental systematic.  We
correspondingly masked out times from BJD 2458488.3 until the end of
the orbit.  The PDC lightcurve initially had 15{,}678 points.  The
quality cut removed 854 points, masking the orbit edges removed an
additional 716, and removing the final few days of orbit 20 removed an
additional 1079.  After cleaning, 83\% of the initial flux
measurements remained.

We normalized these points by dividing out the median flux. We then
subtracted by unity to simplify subsequent analysis.  Many of these
and subsequent processing steps were performed using
\texttt{astrobase}~\citep{bhatti_astrobase_2018}.

@@

Between July 25, 2014 and May 23, 2016, Gaia measured about 300
billion centroid positions of 1{.}6 billion stars
\citep{gaia_collaboration_gaia_2016,lindegren_gaiasoln_2018,gaia_collaboration_gaia_2018}.
These CCD observations were used to estimate positions, proper
motions, and parallax for the brighest 1{.}3 billion stars, including
\ptfo.  There were 121 ``good'' observations of \ptfo, {\it i.e.},
observations that were not strongly downweighted in the astrometric
solution of the source.  The star was assigned a Gaia DR2 identifier
of 3222255959210123904.  It photometric brightness was measured using
selected bands ($G$, $Rp$, and $Bp$) of the Gaia Radial Velocity
Spectrometer \citep{cropper_gaia_2018,evans_gaia_2018}.

%FIXME FROM HERE

One point of note:
Gaia DR2 measured a significant ``astrometric excess'', at a level of
10.3$\sigma$.
This astrometric excess indicates the degree
to which a single-source model fails to explain the observed astrometric
measurements (CITE).

\subsection{Cluster Membership}
The Orion molecular cloud complex has numerous subgroups, with ages
spanning 1 to 15$\,$Myr (CITE).
\ptfo\ has been known to be a member of the Orion OB1a association
since at least CITE (Briceno 2005).
Its relation to the broader Orion complex has been further explored by
CITEX, CITEY, and CITEZ (Briceno 05, 08, 18, whomever Van Eyken cites,
and Kounkel 18).

@@

\section{Gaia Analysis \& Photometric Binarity}
\label{sec:gaia}

% \begin{figure*}[t]
% 	\begin{center}
% 		\leavevmode
% 		\includegraphics[width=0.9\textwidth]{f6.png}
% 	\end{center}
% 	\vspace{-0.7cm}
% 	\caption{ {\bf Positions, kinematics, and HR diagram of \ptfo.}
%   Members of the 25$\,$Ori-1 subgroup are shown with black circles,
%   and were identified by \citet{kounkel_apogee2_2018} through
%   clustering on Gaia-DR2 and APOGEE data.  The ``neighborhood'' (gray
%   circles) is defined as the group of at most $10^4$ randomly selected
%   non-member stars within 5 standard deviations of the mean right
%   ascension, declination, and parallax.
% 	\label{fig:gaia}
% 	}
% \end{figure*}
\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.8\textwidth]{f5.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Hertzsprung-Russell diagram of \ptfo\ and late-type
  members of 25$\,$Ori-1.}
  Members of the 25$\,$Ori-1 group (black circles) were identified by
  \citet{kounkel_apogee2_2018} through clustering on six-dimensional
  Gaia-DR2 and APOGEE-2 data.  The reference ``neighborhood'' (gray
  circles) is the group of at most $10^4$ randomly selected non-member
  stars within 5 standard deviations of the mean 25$\,$Ori-1 right
  ascension, declination, and parallax.  It contains members of the
  full Orion complex with a wider spread of ages, in addition to field
  interlopers.  $G$ is the Gaia broadband, $Bp$ is Gaia blue, $Rp$ is
  Gaia red, and $\omega_{\rm as}$ is the parallax in arcseconds.  The
  $x$-axis limits have been set to show only K and M dwarf members, to
  accentuate \ptfo's separation from the single-star sequence.
	\label{fig:gaia}
	}
\end{figure*}

@@

To assess the cluster membership and potential binarity of \ptfo, we
needed to identify stars with which it was and was not coeval.  The
simplest way to do this---clustering based on six-dimensional position
and kinematic information---had already been done by
\citet{kounkel_apogee2_2018}.  For simplicity, we considered the
members they identified brighter than $G_{\rm Rp}$ of 16.  This
yielded 149 stars in 25$\,$Ori-1, mostly M dwarfs.
\citet{kounkel_apogee2_2018} identified seven other smaller groups in
the Orion complex near the Be star 25$\,$Ori. These groups received
higher numbers, {\it e.g.}, 25$\,$Ori-2.

@@

>>>>

\documentclass[12pt,twocolumn,tighten]{aastex62}
%\documentclass[12pt,twocolumn,tighten,trackchanges]{aastex62}
\usepackage{amsmath,amstext,amssymb}
\usepackage[T1]{fontenc}
\usepackage{apjfonts}
\usepackage[figure,figure*]{hypcap}
\usepackage{graphics,graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage[caption=false]{subfig} % for subfloat

@@

%\objectname{PTFO 8-8695}
%\objectname{PTFO 8-8695b}
%\objectname{CVSO 30}

@@

\begin{abstract}
  PTFO$\,$8-8695b is a candidate hot Jupiter in the 7--10 million year
  old Orion-OB1a cluster. We inspected data from TESS and Gaia to
  clarify whether it is truly a planet.  The Gaia data show that
  PTFO$\,$8-8695 is a photometric binary with respect to members of
  its kinematic group.  The TESS lightcurve shows that the dominant
  variability in this system is a sinusoid with a ``long'' period
  $P_{\rm \ell}=11.96\,$hr, presumably caused by stellar rotation.
  Also present is a complex signal, previously identified as the
  planet candidate, that repeats with a ``short'' period $P_{\rm s}=
  10.74\,$hr.  The two signals beat every 4.48 days.  Although there
  is a dip in the short-period signal, ground-based photometry from
  the past decade shows that the orbital phase of the dip seems to
  have instantaneously jumped, at least once, and perhaps twice.
  Planets do not ``jump'' in orbital phase.  Given the evidence, we
  believe that PTFO$\,$8-8695 is a binary M dwarf in which one star
  shows the long rotation signal, and the other star shows
  ``transient dipping'' that has been seen in a few other young
  weak-lined T Tauri stars.  The origin of these transient dips is
  still undetermined, but our preferred explanation is eclipses of
  clouds of gas or dust at the Keplerian co-rotation radius.
\end{abstract}

@@

Alternatively, between 0.1\% and 1\% of rapidly rotating low-mass
stars in $\mathcal{O}$(10)$\,$Myr old associations show narrow dips in
phase with strong stellar rotation signals \citep{rebull_usco_2018}.
The dips can persist over months, but their depths often vary, and
sometimes change immediately after stellar flares.  The explanation
proposed by \citet{stauffer_orbiting_2017} and
\citet{david_transient_2017} is that a circumstellar cloud of dust or
gas might be orbiting near the Keplerian co-rotation radius.  For this
explanation to be viable for \ptfo, a clear determination of the
stellar rotation period is necessary.  To date this has not been
possible
\citep{van_eyken_ptf_2012,koen_multicolour_2015,raetz_yeti_2016}.

We begin in Section~\ref{sec:observations} by describing newly
available observations from TESS
\citep{ricker_transiting_2015} and Gaia
\citep{gaia_collaboration_gaia_2018}.  The TESS lightcurve shows two
distinct signals, which we extract and analyze in
Section~\ref{sec:tess}.  A long-period sinusoid repeats every 12.0
hours, and is probably stellar rotation.  A short-period dip with an additional
complex modulation repeats every 10.7 hours.  Analyzing the Gaia
data in Section~\ref{sec:gaia}, we show that relative to its kinematic
group, \ptfo\ is a photometric binary.  We collect and discuss the
puzzle pieces in Section~\ref{sec:discussion}.  The orbital phase of
the dip seems to have instantaneously changed over the past decade.
In addition, a number of other young stars show lightcurve
morphologies similar to the short-period signal.  We therefore argue that PTFO$\,$8-8695 is a binary M
dwarf in which one star shows a rotation signal, and the other 
shows ``transient dipping'' caused by eclipses of
material at the Keplerian co-rotation radius.
Section~\ref{sec:conclusions} summarizes our main points.

@@

The 2-minute stacks for \ptfo\ were reduced to lightcurves by the
Science Processing Operations Center (SPOC) at NASA
Ames~\citep{jenkins_tess_2016}.  Our main analysis used the resulting
Presearch Data Conditioning (PDC) lightcurve.  The PDC lightcurve
aperture used pixels chosen to maximize the SNR of the total flux of
the target \citep{smith_kepler_apertures_2017}.  Non-astrophysical
variability was removed through the methods discussed by
\citet{smith_kepler_PDC_2017}.

As an independent check on the shorter cadence SPOC light-curve, we
separately processed the 30-minute image stacks as part of the Cluster
Difference Imaging Photometric Survey (CDIPS;
\citealt{bouma_cluster_2019}).  Our CDIPS lightcurve  of choice used a circular
aperture with radius 1 pixel.

To clean the data, we removed all points with non-zero quality flags
\citep[{\it e.g.},][]{tess_data_product_description_2018}.  We also
masked out the first and last 6 hours of each orbit, since there is
often systematic red noise during those times.  Both the CDIPS and PDC
lightcurves showed a clear discontinuous ``jump'' in the last few days
of orbit 20, which seemed likely to be an instrumental systematic.  We
correspondingly masked out times from BJD 2458488.3 until the end of
the orbit.  The PDC lightcurve initially had 15{,}678 points.  The
quality-flag cut removed 854 points; masking the orbit edges removed an
additional 716; removing the final few days of orbit 20 removed an
additional 1079.  After cleaning, 83\% of the initial flux
measurements remained.

We normalized these points by dividing out the median flux. We opted
to then subtract by unity to simplify subsequent interpretation.  Many
of these and subsequent processing steps were performed using
\texttt{astrobase}~\citep{bhatti_astrobase_2018}.

@@

\subsubsection{Astrometry of \ptfo}

Between July 25, 2014 and May 23, 2016, Gaia measured about 300
billion centroid positions of 1{.}6 billion stars
\citep{gaia_collaboration_gaia_2016,lindegren_gaiasoln_2018,gaia_collaboration_gaia_2018}.
In the Gaia second data release (DR2), these CCD observations were
then used to estimate positions, proper motions, and parallax for the
brighest 1{.}3 billion stars, including \ptfo
\citep{lindegren_gaiasoln_2018}.  There were 121 ``good'' observations
of \ptfo, that is observations that were not strongly down-weighted in
its astrometric solution.  \ptfo\ was assigned a Gaia DR2 identifier
of 3222255959210123904.  It photometric brightness was measured using
selected bands ($G$, $Rp$, and $Bp$) of the Gaia Radial Velocity
Spectrometer \citep{cropper_gaia_2018,evans_gaia_2018}.  We accessed
the pipeline parameters for \ptfo\ using the Gaia
archive\footnote{\url{gea.esac.esa.int/archive/}}.

The majority of Gaia's derived parameters for \ptfo\ agreed with
expectation from former studies
\citep{briceno_cida_2005,van_eyken_ptf_2012}.  One novelty however was
that Gaia DR2 detected a significant ``astrometric excess'', at a
level of 10.3$\sigma$.  
We comment on the significance and interpretation of this
excess in Section~\ref{sec:gaia}.

\subsubsection{Cluster Membership}

The Orion molecular cloud complex has numerous subgroups, with ages
spanning 1 to 15$\,$Myr (CITE).
\ptfo\ has been known to be a member of the Orion OB1a association
since at least CITE (Briceno 2005).
Its relation to the broader Orion complex has been further explored by
CITEX, CITEY, and CITEZ (Briceno 05, 08, 18, whomever Van Eyken cites,
and Kounkel 18).

@@

\section{Gaia Analysis}
\label{sec:gaia}

% \begin{figure*}[t]
% 	\begin{center}
% 		\leavevmode
% 		\includegraphics[width=0.9\textwidth]{f6.png}
% 	\end{center}
% 	\vspace{-0.7cm}
% 	\caption{ {\bf Positions, kinematics, and HR diagram of \ptfo.}
%   Members of the 25$\,$Ori-1 subgroup are shown with black circles,
%   and were identified by \citet{kounkel_apogee2_2018} through
%   clustering on Gaia-DR2 and APOGEE data.  The ``neighborhood'' (gray
%   circles) is defined as the group of at most $10^4$ randomly selected
%   non-member stars within 5 standard deviations of the mean right
%   ascension, declination, and parallax.
% 	\label{fig:gaia}
% 	}
% \end{figure*}
\begin{figure*}[t]
	\begin{center}
		\leavevmode
		%\includegraphics[width=0.8\textwidth]{f5a.pdf}
    \subfloat{
        \includegraphics[width=0.7\textwidth]{f5a}
    }

	\vspace{-0.7cm}
    \subfloat{
        \includegraphics[width=0.7\textwidth]{f5b}
    }
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Evidence for binarity in \ptfo}.
  {\it Top: Hertzsprung-Russell diagram of \ptfo\ and late-type
  members of 25$\,$Ori-1.}
  Members of the 25$\,$Ori-1 group (black circles) were identified by
  \citet{kounkel_apogee2_2018} through clustering on six-dimensional
  Gaia-DR2 and APOGEE-2 data.  The ``neighborhood'' (gray
  circles) is non-member
  stars within 5 standard deviations of the mean 25$\,$Ori-1 right
  ascension, declination, and parallax.  It contains members of the
  Orion complex with its full spread of ages, in addition to field
  interlopers.  $G$ denotes Gaia broadband magnitudes, $Bp$ Gaia blue, $Rp$
  Gaia red, and $\omega_{\rm as}$ the parallax in arcseconds.  The
  $x$-axis limits have been set to show only K and M dwarf members, to
  accentuate \ptfo's separation from the single-star sequence.
  {\it Bottom: Astrometric goodness-of-fit versus $Rp$ magnitude
  for 25$\,$Ori-1 members}.
  The single-source astrometric model provides a poor fit, which
  could be due to stellar variability or binarity.
	\label{fig:gaia}
	}
\end{figure*}

@@

\subsection{Photometric Binarity}
To assess the cluster membership and potential binarity of \ptfo, we
needed to identify stars with which it was and was not coeval.  The
simplest way to do this---clustering based on six-dimensional position
and kinematic information---had already been done by
\citet{kounkel_apogee2_2018}.  For simplicity, we considered the
members they identified brighter than $G_{\rm Rp}$ of 16.  This
yielded 149 stars in 25$\,$Ori-1, mostly M dwarfs.
\citet{kounkel_apogee2_2018} identified seven other smaller groups in
the Orion complex near the Be star 25$\,$Ori. These groups received
higher numbers, {\it e.g.}, 25$\,$Ori-2.

@@

\subsection{Astrometric Binarity}

As noted in Section~\ref{sec:observations}, the Gaia DR2 solution for
\ptfo\ shows a 10.3$\sigma$ astrometric excess.  This astrometric
excess indicates the degree to which a single-source model fails to
explain the observed astrometric measurements.  Specifically, the
single-source astrometric model for yielded $\chi^2=325.2$.  There
are 121 astrometric measurements, and 5 free parameters, and therefore
116 degrees of freedom. The reduced $\chi^2$ is 2.80.  The
majority of stars with comparable brightness in Gaia do not show such
poor goodness-of-fit \citep[][Appendix A]{lindegren_gaiasoln_2018}.

Potential explanations for the poor astrometric fit include photometric
variability and unresolved stellar binarity \citep[{\it
e.g.},][]{rizzuto_ZEIT8_2018,belokurov_unresolved_2020}.
If photometric variability were the cause, we would expect comparably
faint stars in the same kinematic group of Orion to
show similar astrometric excesses, as the majority of
young stars are highly variable.

Using the same 149 members in the 25$\,$Ori-1 subgroup from
\citet{kounkel_apogee2_2018}, we calculated the reduced $chi^2$ for
each member.  The lower panel of Figure~\ref{fig:gaia} shows the
resulting values, as a function of stellar brightness.  \ptfo\ is in
the upper 90$^{\rm th}$ percentile of stars showing astrometric
excesses within the 25$\,$Ori-1 group.  Although we will have to wait
for the full release of the nominal Gaia mission to definitively
determine whether the astrometric excess is caused by stellar binarity
or photometric variability, this result suggests that stellar binarity
is the root cause, because other member stars have identical variability
characteristics but do not show such a large astrometric excess.
%FIXME: you might want to color the points by their TESS amplitude, or
%similar.
% or even better... just do a CUT on TESS amplitude. Require it to be
% like 10% or above -- so LARGER amplitudes than PTFO! That's the best
% test.
